Supporting Information:\\ NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PS lipids

### 2019-03-20 Manuscript/manuscriptPS.tex
Update figures and text related to the experimental data based on emails by Tiago Ferreira

\emph{Numerical simulations of S-DROSS curves.}
The numerical simulations of S-DROSS curves were performed using the SIMPSON simulation package \cite{bak00}
by inputing the $^{13}$C--$^1$H
dipolar couplings, either as determined by the R-PDLF experiments, or as calculated from the known $^2$H quadrupolar couplings \cite{browning80}.
The chemical shift anisotropy and homonuclear couplings were neglected, and the input file {\it{rep2000}}\todo{citation needed?} was used to simulate a random
distribution of bilayer orientations in the samples studied.

\emph{Sample preparation.}
The sample was prepared simply by mixing POPS (1-palmitoyl-2-oleoyl-{\it sn}-glycero-3-phospho-L-serine, purchased from Avanti Polar Lipids
as sodium salt) with water (lipid:water 60:40\,wt-\%) in an Eppendorf tube by mixing and centrifuging the sample repeatedly until a homogeneous viscous fluid was obtained
\todo{How long and how many repeats did this take approximately? Was the homogeneity determined visually?}.
Then 20 mg of the sample was transferred to an NMR insert suitable for 4 mm NMR rotors.

@@

\begin{figure}[!tb]
  \centering
  \includegraphics[width=8.5cm]{../Figs/fig1_POPS.pdf}
  \caption{\label{PShgSIGNSsimpson}
    The headgroup and glycerol backbone region of the (A) INEPT spectrum and
    (B) 2D R-PDLF spectra.
    (C) Experimental S-DROSS data (points), and SIMPSON simulations (blue lines) with
    the C--H bond order parameter values of -0.12 for the $\beta$-carbon, and +0.09 and -0.02
    for the $\alpha$-carbon.
    Dashed gray line is the S-DROSS curve from a SIMPSON simulation with a positive value (+0.02) 
    for the smaller $\alpha$-carbon C--H bond order parameter.
  }
  \todo{I think that the peak labeling would be good to show also in (A).
  There is possibly a new version of (C) coming.}\\
\end{figure}
\begin{figure}[!htb]
  \centering
  \includegraphics[width=9.0cm]{../Figs/PCPScomp.eps}
  \caption{\label{HGorderParameters}
    (A) Chemical structures and labels for the headgroup and glycerol backbone carbons.
    (B) Headgroup and glycerol backbone order parameters of POPS (T = 298 K) measured in this work compared
    with the previously published values from DOPS (T = 303 K, $^2$H NMR, 0.1M of NaCl) \cite{browning80} and 
    POPC  (T = 300 K, $^{13}$C NMR) \cite{ferreira13} experiments. Signs of the PS order parameters
    are measured in this work whereas signs of the PC order parameters are measured previously~\cite{ferreira16}.
    The size of errorbars ($\pm$0.02) shown for $^{13}$C NMR data is justified previously~\cite{botan15,ollila16}. 
  }
  \todo{Replacing the "literature" with "from Ref X" has been suggested  https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/34.
  We can do this just before submission when citation numbers will not change anymore.}
\end{figure}

@@

Since the R-PDLF and previous $^2$H\,NMR experiments \cite{browning80,roux91} give 
only the absolute values of order parameters, we determined the signs of the PS headgroup
order parameters using the S-DROSS experiment \cite{gross97}.
For a given carbon, its S-DROSS dipolar modulation profile in the indirect dimension is a superposition
of sinusoidal functions from the possible orientations of crystallites in the sample (or bilayer patches).
We phase corrected the 2D spectrum in the direct dimension such that positive and negative signs for the C—H
bond order parameter give rise to profiles that initially increase and decrease, respectively.
In practice, it is useful to use the known negative sign of the acyl chain carbons as a reference to perform
the phase correction and interpret the distinct initial slopes of the S-DROSS profiles as shown in figure 1. 
The S-DROSS slice for the $\beta$-carbon clearly shows an initial decrease and therefore its order parameter must be negative.
For the $\alpha$-carbon such analysis is not as trivial due to the two inequivalent order parameters of the two distinct C—H bonds.
However, the beginning of its S-DROSS slice suggests that the larger order parameter of the alpha-carbon is positive and the
decrease towards negative values at longer $t_1$ suggests that the smaller order parameter is negative.    
%
%The S-DROSS slice clearly shows that the order parameter of
%the $\beta$-carbon is negative (Fig. \ref{PShgSIGNSsimpson} C)), \todo{
%Could we explain in a few words how this is clearly seen?}
%which is confirmed by SIMPSON simulations. The beginning of the S-DROSS slice
%suggests that the larger order parameter of the $\alpha$-carbon 
%is positive and the deviation towards negative values with longer T$_1$ times suggests
%that the smaller order parameter is negative.
%
This is confirmed by a SIMPSON simulation
using the value of -0.02 from $^2$H\,NMR experiment \cite{roux91} for the smaller order parameter.
The literature value was used because the
resolution of our R-PDLF experiment was not sufficient to determine the
small value of the order parameter.
The S-DROSS curve from SIMPSON simulation with a positive value for the smaller order parameter
(dashed grey in Fig. \ref{PShgSIGNSsimpson} C)) did not agree with the experiment, 
corroborating the interpretation that the smaller order parameter is negative.

>>>>

\emph{Numerical simulations of S-DROSS curves.}
The numerical simulations of S-DROSS curves were performed using the SIMPSON simulation package \cite{bak00}
by inputing the $^{13}$C--$^1$H
dipolar couplings, either as determined by the R-PDLF experiments, or as calculated from the known $^2$H quadrupolar couplings \cite{browning80}.
The chemical shift anisotropy and homonuclear couplings were neglected, and the SIMPSON input file {\it{rep2000}} was used to simulate a random
distribution of bilayer orientations in the samples studied.

\emph{Sample preparation.}
The sample was prepared simply by mixing POPS powder (1-palmitoyl-2-oleoyl-{\it sn}-glycero-3-phospho-L-serine, purchased from Avanti Polar Lipids
as sodium salt) with water (lipid:water 60:40\,wt-\%) in an Eppendorf tube by mixing and centrifuging the sample approximately 5 to 6 times
until a homogeneous viscous fluid was visually observed. Then 20 mg of the sample was transferred to an NMR insert suitable for 4 mm NMR rotors.

@@

\begin{figure}[!tb]
  \centering
  \includegraphics[width=8.5cm]{../Figs/fig1_POPS.pdf}
  \caption{\label{PShgSIGNSsimpson}
    The headgroup and glycerol backbone region of the (A) INEPT spectrum and
    (B) 2D R-PDLF spectra.
    (C) Experimental S-DROSS data (points), and SIMPSON simulations (blue lines) with
    the C--H bond order parameter values of -0.12 for the $\beta$-carbon, and +0.09 and -0.02
    for the $\alpha$-carbon.
    Dashed gray line is the S-DROSS curve from a SIMPSON simulation with a positive value (+0.02) 
    for the smaller $\alpha$-carbon C--H bond order parameter.
  }
\end{figure}
\begin{figure}[!htb]
  \centering
  \includegraphics[width=9.0cm]{../Figs/PCPScomp.eps}
  \caption{\label{HGorderParameters}
    (A) Chemical structures and labels for the headgroup and glycerol backbone carbons.
    (B) Headgroup and glycerol backbone order parameters of POPS (T = 298 K) measured in this work compared
    with the previously published values from DOPS (T = 303 K, $^2$H NMR, 0.1M of NaCl) \cite{browning80} and 
    POPC  (T = 300 K, $^{13}$C NMR) \cite{ferreira13} experiments. Signs of the PS order parameters
    are measured in this work whereas signs of the PC order parameters are measured previously~\cite{ferreira16}.
    The size of errorbars ($\pm$0.02) shown for $^{13}$C NMR data is justified previously~\cite{botan15,ollila16}. 
  }
  \todo{Replacing the "literature" with "from Ref X" has been suggested  https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/34.
  We can do this just before submission when citation numbers will not change anymore.}
\end{figure}

@@

Since the R-PDLF and previous $^2$H\,NMR experiments \cite{browning80,roux91} give 
only the absolute values of order parameters, we determined the signs of the PS headgroup
order parameters using the S-DROSS experiment \cite{gross97}.
For a given carbon, its S-DROSS dipolar modulation profile in the indirect dimension is a superposition
of sinusoidal functions from the possible orientations of crystallites in the sample (or bilayer patches).
We phase corrected the 2D spectrum in the direct dimension such that positive and negative signs for the C—H
bond order parameter give rise to profiles that initially increase and decrease, respectively.
In practice, we use the known negative sign of the acyl chain carbons as a reference to perform
the phase correction and interpret the distinct initial slopes of the S-DROSS profiles (Fig. \ref{DPslices}). 
The S-DROSS slice for the $\beta$-carbon clearly shows an initial decrease and therefore its order parameter must be negative.
For the $\alpha$-carbon such analysis is not as trivial due to the two inequivalent order parameters of the two distinct C—H bonds.
However, the beginning of its S-DROSS slice suggests that the larger order parameter of the alpha-carbon is positive and the
decrease towards negative values at longer $t_1$ suggests that the smaller order parameter is negative.    
%
%The S-DROSS slice clearly shows that the order parameter of
%the $\beta$-carbon is negative (Fig. \ref{PShgSIGNSsimpson} C)), \todo{
%Could we explain in a few words how this is clearly seen?}
%which is confirmed by SIMPSON simulations. The beginning of the S-DROSS slice
%suggests that the larger order parameter of the $\alpha$-carbon 
%is positive and the deviation towards negative values with longer T$_1$ times suggests
%that the smaller order parameter is negative.
%
This is confirmed by a SIMPSON simulation
using the order parameter values of +0.09 from dipolar coupling measured here (Fig. \ref{DPslices})
and -0.02 from the $^2$H\,NMR experiment~\cite{roux91}.
The literature value for the smaller order parameter was used because the
resolution of our R-PDLF experiment was not sufficient to determine the
small value of the order parameter.
The S-DROSS curve from SIMPSON simulation with a positive value for the smaller order parameter
(dashed grey in Fig. \ref{PShgSIGNSsimpson} C)) did not agree with the experiment, 
corroborating the interpretation that the smaller order parameter is negative.

### 2019-03-20 Manuscript/manuscriptPSsuppl.tex
Update figures by Pavel Buslaev http://nmrlipids.blogspot.com/2018/12/nmrlipids-iv-toward-submission-of.html?showComment=1552649629798#c3793703148326511254
Update figures and text related to the experimental data based on emails by Tiago Ferreira

\pagebreak
\section{Dipolar slices of the R-PDFL experiment}

Slices of the R-PDFL spectra (Fig. \ref{DPslices}) 
show a single splitting for the $\beta$-carbon with the order parameter value of 0.11,
and a superposition of a large and a very small splitting for the $\alpha$-carbon.
The larger splitting from the $\alpha$-carbon gives a order parameter value of 0.09, while the numerical value
from the small splitting cannot resolved with the available resolution.
\begin{figure}[]
%  \centering
  \includegraphics[width=20.0cm]{../Figs/SI_slices.pdf}
  \caption{\label{DPslices}
    Dipolar coupling slices from the R-PDFL experiment.
  }
  \todo{Labeling of carbons and explanations of numbers (chemical shift and order parameter) would be good.
    The splitting corresponding the larger $\alpha$-carbon order parameter should be also shown.
  Also, the size of the figure in the file may not be optimal.}
\end{figure}

@@

\begin{figure*}[]
  \centering
  \includegraphics[width=16.0cm]{../Figs/figS7PC.png}
  \caption{\label{dihedralsHGpc}
    Dihedral angle distributions of the headgroup regions from CHARMM36 and Slipids simulations
    compared between the POPC and POPS lipids. 
    The CHARMM36 POPC simulation is from Ref. \citenum{POPCcharmm36T303K} and Slipids POPC from Ref. \citenum{POPCslipids298K}.
  }
  \todo{Colours with more contrast would be better. The y-axis scale could be zoomed.}
\end{figure*}

\begin{figure*}[]
  \centering
  \includegraphics[width=15.0cm]{../Figs/figS6PC.png}
  \caption{\label{dihedralsGLYpc}
    Dihedral angle distributions of the glycerol backbone regions from CHARMM36 and Slipids simulations
    compared between the POPC and POPS lipids.
    The CHARMM36 POPC simulation is from Ref. \citenum{POPCcharmm36T303K} and Slipids POPC from Ref. \citenum{POPCslipids298K}.
  }
  \todo{Colours with more contrast would be better. The y-axis scale could be zoomed.}
\end{figure*}

>>>>

\pagebreak
\section{Spectral slices in the indirect dimension from the R-PDLF and SDROSS experiments}

The C--H bond order parameters of the headgroup and glycerol backbone carbons are determined 
as $S_{\rm CH}=\Delta\nu/(0.315\times21.5  {\rm kHz})$, where $\Delta\nu$ is the dipolar splitting
given by the largest peak widths observed in the second dimension of the R-PDLF spectra
(blue lines in Fig. \ref{DPslices}), 0.315 is the scaling factor of the R18 recoupling sequence 
and 21.5 kHz is the maximum $^1$H-$^{13}$C dipolar coupling for a C--H bond \cite{dvinskikh04}.
%Slices of the R-PDFL spectra (Fig. \ref{DPslices}) 
%show a single splitting for the $\beta$-carbon with the order parameter value of 0.11,
%and a superposition of a large and a very small splitting for the $\alpha$-carbon.
%The larger splitting from the $\alpha$-carbon gives a order parameter value of 0.09, while the numerical value
%from the small splitting cannot resolved with the available resolution.
The resulting order parameters are in good agreement with the previously reported values
from $^2$H NMR experiments \cite{browning80} (Fig. \ref{HGorderParameters} in the main text).
However, the resolution in our $^{13}$C NMR experiment was not sufficient to detect the
the splitting related to the smaller order parameter of the C--H bond in the $\alpha$ carbon
observed in $^2$H NMR experiments \cite{browning80}. Therefore, the value of  0 $\pm$ 0.02 from our
$^{13}$C NMR experiments is shown figure \ref{HGorderParameters} 
and the magnitude of 0.02 from the literature is used in the SIMPSON calculations in the main text.

Interpretation of the order parameter signs of the $\alpha$ carbon from the SDROSS experiment
is complicated by the presence of distinct order parameters for the two attached hydrogens.
As discussed in the main text, interpretation of the SDROSS curve with the help of SIMPSON
simulations gave values +0.09 and -0.02 for the $\alpha$-carbon order parameters.
To corroborate our interpretation, we measured the SDROSS curve also using the higher 8 kHz
MAS frequence (Fig. \ref{DPslices} bottom), which makes the experiment more sensitive to larger order parameter values.
Also this experiment indicates a postive value for the larger $\alpha$-carbon order parameter.
\begin{figure}[]
%  \centering
  \includegraphics[width=\textwidth]{../Figs/SI_man.pdf}
  \caption{\label{DPslices}
    Spectral slices in indirect dimension from the R-PDLF (top) and S-DROSS (middle and bottom) experiments at distinct chemical shifts.
    The chemical shifts and order parameters (calculated from the dipolar splittings indicated with blue lines in the R-PDLF slices)
    are shown on top each column (chemical shift left, order parameter right).
    The assignment of columns is given in the middle.  
    The SDROSS slices were measured using both 5 KHz (middle) and 8 kHz (bottom) because different MAS frequencies enable the
    higher sensitivity to the order parameters with lower and higher magnitudes, respectively.
    The background noise taken from chemical shift slices without carbon peaks (grey lines in SDROSS figures)
    are used to determine the error bars for $\alpha$ and $\beta$ carbons in figure \ref{PShgSIGNSsimpson} in the main text.
    The SDROSS profile of the acyl chain methyl carbons (left column) was used as reference assuming that these carbons have negative order parameters.  
  }
\end{figure}

@@

\begin{figure*}[]
  \centering
  \includegraphics[width=16.0cm]{../Figs/figS11.png}
  \caption{\label{dihedralsHGpc}
    Dihedral angle distributions of the headgroup regions from CHARMM36 and Slipids simulations
    compared between the POPC and POPS lipids. 
    The CHARMM36 POPC simulation is from Ref. \citenum{POPCcharmm36T303K} and Slipids POPC from Ref. \citenum{POPCslipids298K}.
  }
\end{figure*}

\begin{figure*}[]
  \centering
  \includegraphics[width=15.0cm]{../Figs/figS10.png}
  \caption{\label{dihedralsGLYpc}
    Dihedral angle distributions of the glycerol backbone regions from CHARMM36 and Slipids simulations
    compared between the POPC and POPS lipids.
    The CHARMM36 POPC simulation is from Ref. \citenum{POPCcharmm36T303K} and Slipids POPC from Ref. \citenum{POPCslipids298K}.
  }
\end{figure*}
